---
title: "Modeling Colon Cancer Therapies"
author: "Hayley Zorkic"
date: "2020-11-16"

editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

### 0. Background

From the original authors of the dataset: *"These are data from one of
the first successful trials of adjuvant chemotherapy for colon cancer.
Levamisole is a low-toxicity compound previously used to treat worm
infestations in animals; 5-FU is a moderately toxic (as these things go)
chemotherapy agent. There are two records per person, one for recurrence
and one for death."*

This project will explore the following variables AT THE TIME OF DEATH
from the colon cancer dataset:

-   rx Treatment : Obs(ervation), Lev(amisole), Lev(amisole)+5-FU
-   differ Differentiation of tumour : 1=well, 2=moderate, 3=poor
-   extent Extent of local spread : 1=submucosa, 2=muscle, 3=serosa,
    4=contiguous structures
-   sex (binary) : 1=male - obstruct (binary) obstruction of colon by
    tumour : 1=yes Numeric (more than 10 values)
-   age (numeric) :min years
-   time (numeric) days until death
-   nodes (numeric) number of lymph nodes with detectable cancer

**After omitting NA's, 888 unique patient observations were left.**

```{r}
# load in out libraries
library(tidyverse)
library(survival)

# select just our variables of interest
data <- colon %>% 
  select(id,rx,sex,age,obstruct,nodes,differ, extent, time, etype) %>%
  filter(etype != 1) %>%
  select(-etype) %>% 
  drop_na()
rownames(data) <- NULL
head(data)
```

### 1. MANOVA

A one way MANOVA was conducted to determine the effect of 3 dependent
variables (age, nodes, and time) on two levels of colon obstruction.

```{r}
# select depended variables
dvs <- c("age","nodes","time")

# run a manova
man <- manova(cbind(age,nodes,time)~obstruct, data)
summary(man) #MANOVA 
```

Significant differences were found among the two levels of obstruction
for at least one of the dependent variables, Pillai= 0.020396, pseudo
F()= 6.1351, p=0.0003949.

```{r}
#get univariate ANOVAs from MANOVA object
summary.aov(man) 
```

Univariate ANOVAs for each dependent variable were conducted as
follow-up tests to the MANOVA, using the Bonferroni method for
controlling for Type 1 error rates for multiple comparisons. The
univariate ANOVAs for age and time were also significant F()= 8.0091,
p=0.004759, and F()= 5.8107,p=0.01613.

```{r}
# run our pairwise comparisons
pairwise.t.test(data$time,data$obstruct, p.adj="none")
pairwise.t.test(data$age,data$obstruct, p.adj="none")
```

Post hoc analysis was performed conducting pairwise comparisons to
determine if obstruction levels differed across time and age. There is a
significant difference between obstruction levels in age, but not time
after adjusting for multiple comparisons (bonferroni a = .05/6 =
0.008333333).

The MANOVA assumes the following:

1. Random samples, independent observations
2. Multivariate normality of DVs (or each group)
3. Homogeneity of within-group covariance matrices ANOVA assumes equal
variance for DV within each group MANOVA assumes it for each DV and
equal covariance between any two DVs,
4. Linear relationships among DVs
5. No extreme univariate or multivariate outlines
6. No multicollinearity (i.e., DVs should not be too correlated).

If anything, I believe the time data would violate the normality data because there may be an exponential increase in days until death.

### 2. Randomization test

If we scramble our data up, we break any systematic association between
group and response. On average the groups' mean responses will be the
same. If we do this repeatedly and calculate an F statistic each time,
we get the sampling distribution of F under the Null hypothesis that all
groups have the same mean. We then compare our actual statistic to the
null distribution to see if it is a plausible value in the null
distribution, else all groups do not have the same mean.

```{r}
library(vegan)

#compute distances/dissimilarities
dists <- data%>%select(sex,age,nodes,differ,extent,time)%>%dist

#perform PERMANOVA on distances/dissimilarities
adonis(dists~obstruct,data)
```

```{r}
SST<- sum(dists^2)/888
SSW <- data%>%group_by(obstruct)%>%
  select(obstruct,sex,age,nodes,differ,extent,time) %>%
    do(d=dist(.[-1],"euclidean"))%>%ungroup()%>%
    summarize(sum(d[[1]]^2)/717 + sum(d[[2]]^2)/171)%>%pull

#observed F statistic
F_obs<-((SST-SSW)/1)/(SSW/886) 
F_obs

Fs<-replicate(1000,{
  new <- data%>%mutate(obstruct=sample(obstruct)) #randomly permute response variable (rx)
  SSW <- new%>%
    group_by(obstruct)%>%
    select(obstruct,sex,age,nodes,differ,extent,time) %>%
    do(d=dist(.[-1],"euclidean"))%>%
    ungroup()%>%
    summarize(sum(d[[1]]^2)/717 + sum(d[[2]]^2)/171)%>%
    #calculate new F on randomized data
    pull((SST-SSW)/1)/(SSW/886) 
})

hist(Fs,prob = T)
#Let's look at this distribution! What is the probability of getting a statistic at least as extreme as the one we got in this (null) distribution? That's your p-value!
mean(Fs>F_obs)
```

A PERMANOVA Randomization Test was performed on the data- using distance
matrices for partitioning distance matrices among sources of variation
and fitting linear models (e.g., factors, polynomial regression) to
distance matrices; uses a permutation test with pseudo-F ratios. The p
value generated by the PERMANOVA is p=0.022\<0.05 so there is no
significant difference in F statistics generated under the null
hypothesis that the obstruction groups differ by
sex,age,nodes,differ,extent,and/or time, out F statistic being 5.811.

We performed a by hand PERMANOVA calculation to compare to our actual F
statistic generated in the PERMANOVA from the vegan package and got a p
value of p=0.014\<0.05 so we reject the null hypothesis and can conclude
the obstruction groups differ across the variables.

### 3.Linear Regression Model

A linear regression model with interaction was constructed to predict
obstruction from prescription and number of nodes. All numeric variables
were centered.

```{r}
data$time_c <- data$time - mean(data$time)
data$age_c <- data$age - mean(data$age)
data$nodes_c <- data$nodes - mean(data$nodes)
fit <- lm(time_c ~ rx*nodes_c, data = data)
summary(fit)
```

The following interpretations from the linear regression can be made,
disregarding significance.

-   `Intercept`: -60.04 is the predicted days from the mean time until
    death when rx=Obs and there is an average number of nodes.

-   `rxLev`: Controlling for nodes, the predicted days from the mean
    time until death increase by 13.67 days for Lev prescription users.

-   `rxLev+5U`: Controlling for nodes, the predicted days from the mean
    time until death increase by 175.64 days for Lev+5U prescription
    users.

-   `nodes_c`: Controlling for rx, for every 1 unit increase in nodes
    from the average,the predicted days from the mean time until death
    decrease by 88.72 days.

-   `rxLev:nodes_c`: The slope for nodes_c on time_c is 19.56 higher for
    rxLev users.

-`rxLev+5U:nodes_c`: The slope for nodes\_ on time_c is 26.44 higher
for rxLEv+5U users.

```{r}
data %>% ggplot(aes(nodes_c,time_c,color=rx))+geom_point()+geom_smooth(method="lm", se=F)
```

```{r}
resids<-fit$residuals
fitvals<-fit$fitted.values
plot(fitvals,resids); abline(h=0, col='red')

par(mfrow=c(1,2)); hist(resids); qqnorm(resids); qqline(resids, col='red')

ggplot(data,aes(age_c,time_c,color=rx))+geom_point()

library(sandwich)
library(lmtest)

bptest(fit)
shapiro.test(resids) 
```

According to the plots, the assumptions of linearity appears to be met.
Homoskedasticity had to be confirmed with the the Breusch-Pagan test-
the null hypothesis was rejected (p=0.04951\<0.05) which showed that
homoskedasticity is violated. Normality was formally tested with the
shapiro-wilk test which showed that normality was violated
(p\<0.0000001).

```{r}
coeftest(fit, vcov = vcovHC(fit)) #regression after adjusting standard errors for violation
```

After performing a robust SE calculation with the original linear
regression, the only two significant effects on the time_c are rxLev+5FU
users and an increase in nodes_c- congruent with the original model.

The proportion of variation in the response variable explained by the
model is 10.48%- due to chance, our numbers have some association with
the outcome.

### 4. Bootstrapped SEs of Linear Regression Model

A Bootstrapped SE was performed on our linear regression model for
predicting obstruction from prescription and number of nodes. All
numeric variables were centered.

```{R}
boot_dat<- sample_frac(data, replace=T)
samp_distn<-replicate(5000, {
boot_dat <- sample_frac(data, replace=T) #take bootstrap sample of rows
fit <- lm(time_c~rx*age_c, data=boot_dat) #fit model on bootstrap sample
coef(fit) #save coifs
})
samp_distn %>% t %>% as.data.frame %>% summarize_all(sd)
coeftest(fit)[,1:2]
coeftest(fit, vcov=vcovHC(fit))[,1:2]

```

The original SEs and the robust SEs differ significantly only at
rxLev:nodes_c with robust SE having a greater Std. error than original
SEs. Bootstrapped SEs were greater than Original SEs and robust SEs in
intercept, rxLev, rxLev+5FU but more than half as large than Robust and
Original SEs in age_c, rxLev:age_c, rxLev+5FU:age_c.

### 5. Logistic Regression

A logistic regression was performed without interaction to predict the
binary variable obstruction from rx, age_c, and time_c. The numeric
variables were centered.

```{r}
library(lmtest)

fit2<-glm(obstruct~rx+age_c+time_c, data=data, family="binomial")
coeftest(fit2)
#Can tell three predictors significantly increase the probability of obstruction (significant positive coefficients)
#EXPONENTIATE COEFFICIENTS TO INTERPRET:
#odds scale coefs (multiplicative): these are the ones you interpret!
coef(fit2)%>%exp%>%round(5)%>%data.frame
```

The following interpretations from the logistic regression can be made,
disregarding significance.

-   `Intercept`: Predicted odds of obstruction when rx=Obs, average age,
    and average days until death is 0.24231.

-   `rxLev`: Controlling for age and time, odds of obstruction for Lev
    prescription users is 0.98154 times that the Obs Prescription users.

-   `rxLev+5FU`: Controlling for age and time, odds of obstruction for
    Lev+5FU prescription users is 0.88266 times that the Obs
    Prescription users.

-   `age_c`: Controlling for prescription and time, every year increase
    in age from the mean multiples the odds of obstruction multiplies
    odds by 0.98020.

-   `time_c`: Controlling for prescription and age, every day increase
    in days until death from the mean multiples the odds of obstruction
    multiplies odds by 0.99977.

```{R}
class_diag <- function(probs,truth){
  #CONFUSION MATRIX: CALCULATE ACCURACY, TPR, TNR, PPV
  tab<-table(factor(probs>.5,levels=c("FALSE","TRUE")),truth)
  acc=sum(diag(tab))/sum(tab)
  sens=tab[2,2]/colSums(tab)[2]
  spec=tab[1,1]/colSums(tab)[1]
  ppv=tab[2,2]/rowSums(tab)[2]
  f1=2*(sens*ppv)/(sens+ppv)

  if(is.numeric(truth)==FALSE & is.logical(truth)==FALSE) truth<-as.numeric(truth)-1
  
  #CALCULATE EXACT AUC
  ord<-order(probs, decreasing=TRUE)
  probs <- probs[ord]; truth <- truth[ord]
  
  TPR=cumsum(truth)/max(1,sum(truth)) 
  FPR=cumsum(!truth)/max(1,sum(!truth))
  
  dup<-c(probs[-1]>=probs[-length(probs)], FALSE)
  TPR<-c(0,TPR[!dup],1); FPR<-c(0,FPR[!dup],1)
  n <- length(TPR)
  auc<- sum( ((TPR[-1]+TPR[-n])/2) * (FPR[-1]-FPR[-n]) )

  data.frame(acc,sens,spec,ppv,f1,auc)
}

prob<-predict(fit2,type="response") #get predictions for each patient. prob>.5
class_diag(prob,data$obstruct)


```

From the confusion matrix, we can see the accuracy of the model is
0.8074324, the sensitivity is 0, the specificity is 1, the precision is
NaN because the sensitivity is 0, and the auc is 0.580754... not great.

Below is the density plot of the log odds colored by obstruction binary
outcome.

```{r}
data$logit<-predict(fit2) #get predicted log-odds (logits)

#plot logit scores for each truth category

data %>% mutate(obstruct=factor(obstruct,levels=c("1","0"))) %>% #changing order so color red=malig
ggplot(aes(logit, fill=obstruct))+geom_density(alpha=.3)+
  geom_vline(xintercept=0,lty=2)
```

Below is the ROC curve and AUC calculations.

```{r}
library(pROC)
library(plotROC)#Compare with this AUC calculator!
ROCplot<-ggplot(data)+geom_roc(aes(d=obstruct,m=prob), n.cuts=0) 
ROCplot
calc_auc(ROCplot)
```

The AUC of 0.580754 is Bad! It is hard to determined obstruction from
rx, age_c, and time_c.

### 6. Logistic Regression with ALL variables

A logistic regression was performed without interaction to predict the
binary variable obstruction from all of the variables. The numeric
variables were centered.

```{r}
library(lmtest)
fit3<-glm(obstruct~rx+sex+age_c+differ+extent+time_c+nodes_c, data=data, family="binomial")
coeftest(fit3)
#Can tell three predictors significantly increase the probability of obstruction (significant positive coefficients)
#EXPONENTIATE COEFFICIENTS TO INTERPRET:
#odds scale coefs (multiplicative): these are the ones you interpret!
coef(fit3)%>%exp%>%round(5)%>%data.frame
prob<-predict(fit3,type="response") #get predictions for each patient. prob>.5
class_diag(prob,data$obstruct)
```

From the confusion matrix, the accuracy of the model is 0.8074324, the
sensitivity is 0, the specificity is 1, the precision is NaN because the
sensitivity is 0, and the auc is 0.6179011...Poor.

A 10-fold CV was run with the model.

```{R}
set.seed(1234)
k=10 #choose number of folds

data10<-data[sample(nrow(data)),] #randomly order rows
folds<-cut(seq(1:nrow(data10)),breaks=k,labels=F) #create folds

diags<-NULL
for(i in 1:k){
  ## Create training and test sets
  train<-data10[folds!=i,] 
  test<-data10[folds==i,]
  
  truth<-test$obstruct ## Truth labels for fold i
  
  ## Train model on training set (all but fold i)
  fit<-glm(obstruct~rx+sex+age_c+differ+extent+time_c+nodes_c,data=train,family="binomial")
  
  ## Test model on test set (fold i) 
  probs<-predict(fit,newdata = test,type="response")
  
  ## Get diagnostics for fold i
  diags<-rbind(diags,class_diag(probs,truth))
}

summarize_all(diags,mean) #average diagnostics across all k folds
```

The accuracy of the model is 0.8074183, the sensitivity is 0, the
specificity is 1, the precision is NaN because the sensitivity is 0, and
the auc is 0.5654982...Bad.

Next, a LASSO was run on the logistic regression.

```{r}
library(glmnet)
y<-as.matrix(data$obstruct) #grab response
x<-model.matrix(obstruct~rx+sex+age_c+differ+extent+time_c+nodes_c,data=data)[,-1] #grab predictors
x<-scale(x)
glm(y~x,family=binomial)
cv <- cv.glmnet(x,y, family="binomial") #picks an optimal value for lambda through 10-fold CV

#make a plot of the coefficients for different values of lambda
{plot(cv$glmnet.fit, "lambda", label=TRUE); abline(v = log(cv$lambda.1se)); abline(v = log(cv$lambda.min),lty=2)}

cv<-cv.glmnet(x,y,family="binomial")
lasso<-glmnet(x,y,family="binomial",lambda=cv$lambda.1se)
coef(lasso)
```

The only retained variable after LASSO is age_c.

The logistic regression was rerun using only the LASSO retained variable
age_c.

```{r}
#cross-validating lasso model
set.seed(1234)
k=10

datalasso<-data[sample(nrow(data)),] #randomly order rows
folds<-cut(seq(1:nrow(datalasso)),breaks=k,labels=F) #create folds

diags<-NULL
for(i in 1:k){
  train <- datalasso[folds!=i,] #create training set (all but fold i)
  test <- datalasso[folds==i,] #create test set (just fold i)
  truth <- test$obstruct #save truth labels from fold i
  
  fit <- glm(obstruct~age_c, 
             data=train, family="binomial")
  probs <- predict(fit, newdata=test, type="response")
  
  diags<-rbind(diags,class_diag(probs,truth))
}

diags%>%summarize_all(mean)
```

After performing a 10-fold CV using only the variables lasso selected:
the model's out-of-sample AUC, 0.5527102, was worse than the regression
above, 0.5654982.
